---
title: 'MA 590-S1 HW1'
author: "Dinesh Jaykumar Kodwani"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE,cache=FALSE)
```


```{r,include=FALSE,cache=FALSE}
inpack=installed.packages()
for(pp in c("tidyverse","sandwich","lmtest","splines")){
  if(!is.element(pp,inpack[,1])) install.packages(pp)
  require(pp)
}
```



```{r,message=FALSE,warning=FALSE}
library(tidyverse)
library(sandwich)
library(lmtest)
library(splines)
```

Download `covid.csv` from [here] and save it in the same folder as this file (if you want to save it in a different folder, edit the filepath below.


```{r,message=FALSE,warning=FALSE}
covid=read_csv("covid.csv")
```
Each line in the dataset is a US county.
`morbJul`, `incJul`, and `incAug` give the number of detected covid cases (`incJul` or `incAug`, for "incidence") or deaths (`morbJul` for morbidity) in the county from July or August 2020, divided by the county population. The dataset also includes:

- `masksALWAYS`: The proportion of respondents to a survey who indicated that they "Always" wear masks when in groups outside of the family (or "bubble"), 
- `population`, `landArea`, `density`: The county's population (c. 2018), land area, and density (the ratio of the two)
- `region`, the region of the country 
- `maskOrder`, whether there was some sort of mask mandate in place. 

The goal will be to use OLS (aka multiple linear regression) to model the counties' August 2020 covid cases (`incAug`) as a function of some of the other variables. 

### Part A

Regress `incAug` on  `incJul`, and `region`. Inspect the residuals vs fitted-values plot. Does it look like a good fit?

```{r}
mod0 <- lm(incAug~incJul+region,data=covid)
plot(mod0,which=1)
```
Make histograms of `covid$incAug`, `log(covid$incAug)`, `covid$incJul`, and `log(covid$incJul)`
```{r}
hist(covid$incAug)
hist(log(covid$incAug))
```
Refit the previous model, but this time with `log(incAug)` and `log(incJul)` replacing `incAug` and `incJul`  and check the residual-fitted value plot again. Did the model improve? Does it fit well? 

```{r}
mod1 <- lm(log(incAug)~log(incJul)+region,data=covid)
plot(mod1,which=1)
```

Test whether each of the coefficients is positive or negative (using HC standard errors). Interpret the results--what do each of the coefficients mean? What can we conclude about them based on the hypothesis tests?

Make a scatterplot of `log(incJul)` versus `log(incAug)` with points colored differently by `region`, and super-impose the model fit. 

```{r}
 coeftest(mod1,vcov.=vcovHC)
```

**Answer**

The expected `log(incAug)` for a county in the North Central region with `log(incJul)=0` (which is impossible) is 
`r round(coef(mod1)[1],2)`. 
Given two counties in the same region, but which differed in `log(incJuly)` by 1, we would expect their `log(incAug)` to differ by about 
`r round(coef(mod1)["log(incJul)"],2)`. 
In other words, if `incJuly` in one county was $e=2.7$ times greater than in the second county, we would expect `incAug` to be 
$e^{`r round(coef(mod1)["log(incJul)"],2)`}=$
`r round(exp(coef(mod1)["log(incJul)"]),2)` times greater. 
Compared to counties in the North Central region (the reference category) with the same July incidence, we expect counties in the Northeast to have $ e^{`r round(coef(mod1)["regionNortheast"],2)`}=`r round(exp(coef(mod1)["regionNortheast"]),2)`$ times the incidence, counties in the South to have $e^{`r round(coef(mod1)["regionSouth"],2)`}=`r round( exp(coef(mod1)["regionSouth"]),2)`$ times the incidence incidence, and counties in the West to have $e^{`r round(coef(mod1)["regionNortheast"],2)`}=`r round( exp(coef(mod1)["regionNortheast"]),2)`$ times the incidence incidence. Based on the hypothesis test results, we can be confident that the relationship between July and August incidence is positive and that counties in the Northeast and West had lower August incidence than North Central counties with comparable July incidence. We do not have strong evidence as to whether counties in the South had higher or lower August incidence than North Central counties with comparable July incidence.

**End**

```{r}
pred <- predict(mod1)
ggplot(covid,aes(log(incJul),log(incAug),color=region))+geom_point(alpha=0.2)+
  geom_line(aes(y=pred))
```

**Answer**

The model appears to fit well–the dots are roughly symmetrically scattered around the 0 line. 

**End**

### Part B
You suspect that the relationship between July and August incidence might vary by region--partly because the pandemic evolved differently in different parts of the country, and possibly because counties in different regions may have reacted differently to July data. 

Center `log(julInc)`:
```{r}
covid$logincJulC=log(covid$incJul)-mean(log(covid$incJul))
```
and then add an interaction between `logincJulC` and `region` into the previous model and test whether each of the coefficients is positive or negative (using HC standard errors).  


Model and coefficients:
```{r}
## PUT CODE FOR THE MODEL AND HYPOTHESIS TEST HERE (without the "#")
mod2 <- lm(log(incAug)~logincJulC*region,data=covid)
coeftest(mod2, vconv. = vconvHC)
```

**Answer**

WRITE YOUR ANSWER (TEXT) HERE
The results indicate the differences in how incJul affects incAug in different regions. The expected log(incAug) for a county in the North Central (reference) region is when logincJulC = 0 (at its mean since it is centered) is -5.79. A change of logincJulC by 1, when all else is constant, corresponds to a 0.49 increase in log(incAug). e^0.49 = 1.63, meaning 63% increase in the North Central region. Whereas, when everything is constant except for regionNortheast, log(incAug) is 0.41 lower than the North Central region. e^-0.41 = 0.65, North Central region is 65% of that in the North Central region. logincJulC:regionNortheast being 0.34 suggests that for every unit 1 increase in logincJulC, log(incAug) increases by 0.34 in the Northeast compared to the North Central region. 0.49 + 0.34 = 0.83, for every 1 unit increase in logincJulC, the log(incAug) increases by 0.83 in the North east. 
For the regionWest, when logincJulC = 0, logincAug in West is 0.29 lower than North Central region. e^-0.299 = 0.74, West is 74% of that in North Central region when logincJulC is at mean. In the interaction term, 0.49 + 0.20 = 0.69, for every 1 unit increase in logincJulC results in log(incAug) to increase by 0.69. Very low p-values for Northeast and West regions and their interaction terms suggest strong evidence, whereas a high p-value for regionSouth, suggest no strong evidence. Whereas, in the interaction term, 0.49 + 0.15 = 0.64, tells us that when logincJulC increases by 1 unit, log(incAug) increases by 0.64. 
**End**

Recreate the plots from the previous model (fitted values versus residuals and a scatterplot with model predictions) for the new model with the interaction. Does the model fit well? 

```{r}
## PUT CODE FOR THE PLOTS HERE (without the "#")
plot(mod2, which=1)
pred2 <- predict(mod2)
ggplot(covid,aes(log(incJul),log(incAug),color=region))+geom_point(alpha=0.2)+
  geom_line(aes(y=pred2))
```

**Answer**

WRITE YOUR ANSWER (TEXT) HERE
The model seems to fit well-the dots are roughly scattered around their respective region lines. 

**End**

```{r}
# Extract coefficients from the model
coeffs <- coef(mod2)

# Print the equation in a readable format
cat("log(incAug) =", 
    round(coeffs[1], 4), "+", 
    round(coeffs["logincJulC"], 4), "* logincJulC +",
    round(coeffs["regionNortheast"], 4), "* regionNortheast +",
    round(coeffs["regionSouth"], 4), "* regionSouth +",
    round(coeffs["regionWest"], 4), "* regionWest +",
    round(coeffs["logincJulC:regionNortheast"], 4), "* logincJulC:regionNortheast +",
    round(coeffs["logincJulC:regionSouth"], 4), "* logincJulC:regionSouth +",
    round(coeffs["logincJulC:regionWest"], 4), "* logincJulC:regionWest"
)

```

